Supp mat: time series paper

### 2017-10-01 writing/manuscript_draft.tex
Version submitted to Biological Conservation and the second version to PeerJ preprints

Title: Minimum time required to detect population trends: the need for
long-term monitoring programs \vspace{7 mm}

Author: \textsc{Easton R. White$^{1,2}$} \vspace{3 mm}

Address:\emph{
    \\$^1$Center for Population Biology \\
    University of California, Davis \\
    2320 Storer Hall \\
        University of California, Davis \\
        One Shields Avenue Davis, CA, USA} \vspace{3 mm}

\(^2\)Corresponding author:
\href{mailto:eawhite@ucdavis.edu}{\nolinkurl{eawhite@ucdavis.edu}}
\vspace{3 mm}

Number of words (including 300 words per figure): 5900 \vspace{1 mm}

Number of figures: 3 \vspace{1 mm}

Supplementary material: 11 figures and 1 table \vspace{3 mm}

Data and code for all the figures and tables can be found at
(\url{https://github.com/erwhite1/time-series-project}).

\vspace{3 mm}

\pagebreak

@@

\begin{abstract}
Long-term time series are necessary to better understand population dynamics, assess species' conservation status, and make management decisions. However, population data are often expensive, requiring a lot of time and resources. When is a population time series long enough to address a question of interest? We determine the minimum time series length required to detect significant increases or decreases in population abundance. To address this question, we use simulation methods and examine 878 populations of vertebrate species. Here we show that 15-20 years of continuous monitoring are required in order to achieve a high level of statistical power. For both simulations and the time series data, the minimum time required depends on trend strength, population variability, and temporal autocorrelation. These results point to the importance of sampling populations over long periods of time. We argue that statistical power needs to be considered in monitoring program design and evaluation. Time series less than 15-20 years are likely underpowered and potentially misleading.
\end{abstract}

@@

How long of a time series is actually necessary? This question has
important implications for both research and management (Nichols and
Williams 2006). Scientists need to know the time series length required
to address a specific question. A short time series may lead to wrong
conclusions given large natural year-to-year variability. Managers need
to know when action is needed for a population. Therefore, they must
understand when population trend over time is actually meaningful. The
IUCN Red List Categories and Criteria suggest, under Criterion A2, a
species qualifies as vulnerable if it has experienced a 30\% decline
over 10 years, or 3 generations (IUCN 2012). For both scientific and
management questions, because sampling is typically expensive, we also
do not want to sample for longer than is necessary. For example, Gerber,
DeMaster, and Kareiva (1999) investigated the minimum time series
required to estimate population growth of the endangered, but
recovering, eastern North Pacific gray whale
(\emph{Eschrichtius robustus}). They used a long-term census to
retroactively determine the minimum time series required to assess
threat status. They found that only 11 years were needed, eight years
before the delisting decision was made. This highlights the importance
of estimating the minimum time series required as an earlier decision
would have saved time and money (Gerber, DeMaster, and Kareiva 1999).

@@

For time series data, two general approaches to estimating sample size
are appropriate. Simulations can be designed for a specific population
and question (Bolker 2008; P. C. Johnson et al. 2015). Simple models can
be simulated with parameter values corresponding to a population of
interest (Gerrodette 1987). Statistical power can then be calculated as
the fraction of simulations that meet some criteria. The specific
criteria depend on the question at hand. For example, given a time
series, when is the slope from linear regression significantly different
from zero? In other words, when is the time series significantly
increasing or decreasing? It is then possible to determine how power
changes with a variable of interest. For example, time series can be
simulated for different lengths of time. From these simulations, the
minimum time series length required is calculated to meet certain levels
of statistical significance and power (Bolker 2008).

In addition to using simulations, empirical time series can also be
used. Multiple replicates of similar populations are usually not
available, but it is possible to subsample an empirical time series
(Gerber, DeMaster, and Kareiva 1999). Subsamples of different lengths
can then be evaluated to see which fraction of subsamples meet some
criteria, again a measure of statistical power. Similar to the
simulation approach, these measures of power can be used to determine
the minimum time series required for a particular question of interest.

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time (Wagner,
Vandergoot, and Tyson 2009; Giron-Nava et al. 2017). For example, Rhodes
and Jonzen (2011) examined the optimal allocation of effort between
spatial and temporal replicates. Using simple populations models, they
found that the allocation of effort depends on environmental variation,
spatial and temporal autocorrelation, and observer error. Rueda-Cediel
et al. (2015) also used a modeling approach, but parameterized a model
specifically for a threatened snail, \emph{Tasmaphena lamproides}. They
found that for this short-lived organism, 15 years was adequate to
assess long-term trends in abundance. Hatch (2003) examined seabird
populations and determined the minimum time required to achieve high
statistical power ranged between 11 and 69 years. However, these
studies, and other past work, have typically focused either on
theoretical aspects of monitoring design or focused on only a few
species.

@@

We use a database of 2444 population time series compiled in (Keith et
al. 2015); they compared the predictability of growth rates among
populations. The data are originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010). We filtered out
short time series (less than 35 years), and those with missing data, and
were left with 878 time series. The data includes information on 478
vertebrate species with a focus on mammals, birds, and fish. The data
also includes information on generation length and census
specifications. For each time series, we also calculated variables of
interest like variance in population size, long-term trend in abundance
(slope coefficient from simple linear regression), and temporal
autocorrelation. All analyses were conducted in R (R Core Team 2016).

@@

We assume that each time series in our database is long enough to
include all necessary information (e.g.~variability) about the
population. In other words, each time series is a representative sample.
We first take all possible contiguous subsamples of each time series.
For example, a time series of 35 years would have 34 possible contiguous
subsamples of length 2, 34 possible contiguous subsamples of length 3,
and continuing until 1 possible contiguous subsample of length 35
(Gerber, DeMaster, and Kareiva 1999; Giron-Nava et al. 2017). Next, we
run a linear regression for each subsampled time series. Then, we
determine the fraction of subsamples of a particular length that have
estimated slope coefficients which are statistically different from
zero. We only look at the fraction of samples where the long-term, or
``true'', time series also has a significant slope. This fraction is a
measure of statistical power. Lastly, we determine which subsample
length is required to achieve a certain threshold of statistical power
(0.8, Cohen (1992)). The subsampled length meeting these criteria is the
minimum time series length required (\(T_{min}\)).

@@

We chose a simple model, but any other population model could be used
(see example in Fig. \ref{fig:min_time_dist_log_pop}). Ideally, the
specific model choice should be tailored to the population of interest.
We explored how the simulation approach can be applied to more
biologically-realistic population models (Fig. \ref{fig:shark_example}).
More specifically, we determined the minimum time required to estimate
long-term population trends using a stochastic, age-structured model of
lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
2014). We found that over 27 years of continuous monitoring were needed
in this particular scenario (Fig. \ref{fig:shark_example}). Similar to
the simulation approach described above, the minimum time required for
the lemon shark population was strongly dependent on model parameters.

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Distributions of the minimum time required for four different
biological classes. The minimum time required calculation corresponds to
a significance level of 0.05 and statistical power of
0.8.\label{fig:min_time_dist}}
\end{figure}

@@

In line with theoretical predictions (Rhodes and Jonzen 2011), we also
found \(T_{min}\) was strongly correlated with the trend strength,
variability in population size, and temporal autocorrelation (Fig.
\ref{fig:correlates}). Contrary to our prior expectations, we also found
that \(T_{min}\) did not correlate with any biological variables of
interest (Fig. \ref{fig:biological_correlates}). We initially
hypothesized that species with longer lifespans or generation times may
require a longer sampling period. Our result could have been a result of
at least two factors. First, the data we used may not include a diverse
enough set of species with different life history traits. Second, the
question we poised, whether a population is increasing or decreasing,
was specifically concerned with trends in population density over time.
Therefore, life-history characteristics may be more important for other
questions, like estimating species extinction risk (J. A. Hutchings et
al. 2012).

An important related question, is the optimal allocation of sampling
effort in space versus time. In a theoretical investigation of this
question, Rhodes and Jonzen (2011) found that the optimal allocation of
sampling depended strongly on temporal and spatial autocorrelation. If
spatial population dynamics were highly correlated, then it was better
to sample more temporally, and vice versa. Our work supports this idea
as populations with strong temporal autocorrelation needed less years of
sampling (Fig. \ref{fig:correlates}). Morrison and Hik (2008) also
studied the optimal allocation of sampling effort in space versus time,
but used emprical data from a long-term census of the collared pika
(\emph{Ochotona collaris}) found in the Yukon. They estimated long-term
growth rates among three subpopulations over a 10-year period. They
found that censuses less than 5 years may be misleading and that
extrapolating from one population to another, even when nearby
geographically, may be untenable.

Seavy and Reynolds (2007) asked whether statistical power was even a
useful framework for assessing long-term population trends. They used 24
years of census data on Red-tailed Tropicbirds
(\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50\%
decline over 10 years almost always resulting in high statistical power
(above 0.8). Therefore, they cautioned against only using power analyses
to design monitoring schemes and instead argued for metrics that would
increase precision. For example, Seavy and Reynolds (2007) suggest
improving randomization, reducing bias, and increases detection
probability when designing and evaluating monitoring programs. We agree
that power analyses should not be the only consideration when designing
monitoring schemes. However, unlike Seavy and Reynolds (2007), our
results indicate that longer than 10 years is often needed to achieve
high statistical power. Therefore, the design of monitoring programs
should include calculations of statistical power, the allocation of
sampling in space versus time (Rhodes and Jonzen 2011), and metrics to
increase precision. Ideally, a formal decision analysis to evaluate
these different factors would be conducted to design or assess any
monitoring program (Hauser, Pople, and Possingham 2006; McDonald-Madden
et al. 2010).

@@

Our work has some limitations in determining the minimum time series
length required. First, \(T_{min}\) is particular to the specific
question of interest. An additional complication is that for our
empirical approach, the subsampling of the full time series allows for
estimates of power, but the individual subsamples are clearly not
independent of one another. Further, estimates of \(T_{min}\) depend on
chosen values of \(\alpha\) and \(\beta\) (Fig.
\ref{fig:min_time_vs_alpha_beta}). In an ideal setting, we would build a
specific population model parameterized for each population of interest.
Then, model simulations could be used to estimate the minimum time
series required to address each specific question of interest. Clearly,
this is not always practical, especially if conducting analyses for a
wide array of species as we do here. In addition, our statistical models
suggest that \(T_{min}\) does not correlate with any biological
variables of interest, at least for the question of linear regression
(Fig. \ref{fig:biological_correlates}). Therefore, it is not possible to
use these results to predict \(T_{min}\) for another population, even if
the population is of a species with a similar life-history to one in our
database.

@@

We use a database of 878 populations to determine the minimum time
series length required to detect population trends. This goes beyond
previous work that either focused on theoretical investigations or a
limited number of species. We show that to identify long-term changes in
abundance, 15-20 years of continuous monitoring are often required (Fig.
\ref{fig:min_time_dist}). In line with theoretical predictions
(Gerrodette 1987), we also show that \(T_{min}\) is strongly correlated
with the long-term population trend (i.e.~rate of increase), variability
in population size, and the temporal autocorrelation (Fig.
\ref{fig:correlates}). Our work implies that for many populations, time
series less than 15-20 years are probably not reliable for detecting
population trends. This result highlights the importance of long-term
monitoring programs. From both a scientific and management perspective
estimates of \(T_{min}\) are important. If a time series is too short,
we lack statistical power to reliably detect long-term population
trends. In addition, a time series that is too long may be a poor use of
already limited funds (Gerber, DeMaster, and Kareiva 1999). Future work
should examine other species, with a wider range of life history
characteristics. In addition, similar approaches can be used to
determine the minimum time series length required to address additional
questions of interest.

@@

\hypertarget{refs}{}
\hypertarget{ref-Bolker2008}{}
Bolker, Benjamin M. 2008. \emph{Ecological Models and Data in R}. 1st
ed. Princeton, New Jersey: Princeton University Press.

@@

@@

>>>>

\textbf{Minimum time required to detect population trends: the need for long-term monitoring programs}
\vspace{3 mm}

Easton R. White\(^{*}\)

\begin{footnotesize} Center for Population Biology, University of California, Davis, California 95616 USA

*eawhite@ucdavis.edu
\end{footnotesize}

\vspace{5 mm}

@@

\begin{abstract}
Long-term time series are necessary to better understand population dynamics, assess species' conservation status, and make management decisions. However, population data are often expensive, requiring a lot of time and resources. When is a population time series long enough to address a question of interest? We determine the minimum time series length required to detect significant increases or decreases in population abundance. To address this question, we use simulation methods and examine 878 populations of vertebrate species. Here we show that 15-20 years of continuous monitoring are required in order to achieve a high level of statistical power. For both simulations and the time series data, the minimum time required depends on trend strength, population variability, and temporal autocorrelation. These results point to the importance of sampling populations over long periods of time. We argue that statistical power needs to be considered in monitoring program design and evaluation. Time series less than 15-20 years are likely under-powered and potentially misleading.
\end{abstract}

@@

How long of a time series is actually necessary? This question has
important implications for both research and management (Nichols and
Williams 2006). Scientists need to know the time series length required
to address a specific question. A short time series may lead to wrong
conclusions given large natural year-to-year variability (McCain,
Szewczyk, and Knight 2016). Managers need to know when action is needed
for a population. Therefore, managers must understand when population
trend over time is actually meaningful. For example, the IUCN Red List
Categories and Criteria suggest, under Criterion A2, a species qualifies
as vulnerable if it has experienced a 30\% decline over 10 years, or 3
generations (IUCN 2012). For both scientific and management questions,
because sampling is typically expensive, we also do not want to sample
for longer than is necessary. For example, Gerber, DeMaster, and Kareiva
(1999) investigated the minimum time series required to estimate
population growth of the endangered, but recovering, eastern North
Pacific gray whale (\emph{Eschrichtius robustus}). They used a long-term
census to retroactively determine the minimum time series required to
assess threat status. They found that only 11 years were needed, eight
years before the delisting decision was made. This highlights the
importance of estimating the minimum time series required as an earlier
decision would have saved time and money (Gerber, DeMaster, and Kareiva
1999).

@@

For time series data, two general approaches to estimating sample size
are appropriate. Simulations can be designed for a specific population
and question (Bolker 2008; P. C. Johnson et al. 2015). Simple models can
be simulated with parameter values corresponding to a population of
interest (Gerrodette 1987). Statistical power is the proportion of
simulations that meet some set of criteria. The specific criteria depend
on the question at hand. For example, given a time series, when is the
slope from linear regression significantly different from zero? In other
words, when is the time series significantly increasing or decreasing?
It is then possible to determine how power changes with a variable of
interest. For example, time series can be simulated for different
lengths of time. From these simulations, the minimum time series length
required to meet certain levels of statistical significance and power is
estimated (Bolker 2008).

In addition to using simulations, empirical time series can also be
used. Multiple replicates of similar populations are usually not
available, but it is possible to subsample an empirical time series
(Gerber, DeMaster, and Kareiva 1999; Brashares and Sam 2005). Subsamples
of different lengths can then be evaluated to estimate the proportion of
subsamples meeting some criteria, again a measure of statistical power.
Similar to the simulation approach, this measures of power can be used
to determine the minimum time series required for a particular question
of interest.

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time (Wagner,
Vandergoot, and Tyson 2009; Giron-Nava et al. 2017). For example, Rhodes
and Jonzen (2011) examined the optimal allocation of effort between
spatial and temporal replicates. Using simple populations models, they
found that the allocation of effort depends on environmental variation,
spatial and temporal autocorrelation, and observer error. Rueda-Cediel
et al. (2015) also used a modeling approach, but parameterized a model
specific for a threatened snail, \emph{Tasmaphena lamproides}. They
found that for this short-lived organism, 15 years was adequate to
assess long-term trends in abundance. Hatch (2003) examined seabird
populations and determined the minimum time required to achieve high
statistical power ranged between 11 and 69 years. However, these
studies, and other past work, have typically focused either on
theoretical aspects of monitoring design or focused on only a few
species.

@@

We use a database of 2444 population time series compiled in (Keith et
al. 2015); they compared the predictability of growth rates among
populations. The data are originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010). We filtered out
short time series (less than 35 years), and those with missing data, and
were left with 878 time series. The data includes information on 478
vertebrate species with a focus on mammals, birds, and fish. The data
also includes information on generation length and survey
specifications. For each time series, we also calculated variables of
interest like variance in population size, long-term trend in abundance
(slope coefficient from simple linear regression), and temporal
autocorrelation. All analyses were conducted in R (R Core Team 2016).

@@

We assume that each time series in our database is long enough to
include all necessary information (e.g.~variability) about the
population. In other words, each time series is a representative sample.
We first take all possible contiguous subsamples of each time series.
For example, a time series of 35 years would have 34 possible contiguous
subsamples of length 2, 34 possible contiguous subsamples of length 3,
and continuing until 1 possible contiguous subsample of length 35
(Gerber, DeMaster, and Kareiva 1999; Giron-Nava et al. 2017). Next, we
run a linear regression for each subsampled time series. Then, we
determine the proportion of subsamples of a particular length that have
estimated slope coefficients which are statistically different from
zero. We only look at the proportion of samples where the long-term, or
``true'', time series also has a significant slope. This proportion is a
measure of statistical power. Lastly, we determine which subsample
length is required to achieve a certain threshold of statistical power
(0.8, Cohen (1992)). The minimum subsampled length that met these
criteria is the minimum time series length required (\(T_{min}\)).

@@

We chose a simple model, but any other population model could be used
(see example in Fig. \ref{fig:min_time_dist_log_pop}). Ideally, the
specific model choice should be tailored to the population of interest.
We explored how the simulation approach could be applied to more
biologically-realistic population models (Fig. \ref{fig:shark_example}).
More specifically, we determined the minimum time required to estimate
long-term population trends using a stochastic, age-structured model of
lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
2014). We found that over 27 years of continuous monitoring were needed
in this particular scenario (Fig. \ref{fig:shark_example}). Similar to
the simulation approach described above, the minimum time required for
the lemon shark population was strongly dependent on model parameters.

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Distributions of the minimum time required for populations from
four different biological classes. The minimum time required calculation
corresponds to a significance level of 0.05 and statistical power of
0.8.\label{fig:min_time_dist}}
\end{figure}

@@

In line with theoretical predictions (Rhodes and Jonzen 2011), we also
found \(T_{min}\) was strongly correlated with the trend strength,
variability in population size, and temporal autocorrelation (Fig.
\ref{fig:correlates}). Contrary to our prior expectations, we also found
that \(T_{min}\) did not correlate with any biological variables of
interest (Fig. \ref{fig:biological_correlates}). We initially
hypothesized that species with longer lifespans or generation times may
require a longer sampling period. Our result could have been a result of
at least two factors. First, the data we used may not include a diverse
enough set of species with different life history traits. Second, the
question we poised, whether a population is increasing or decreasing,
was specifically concerned with trends in population density over time.
Therefore, life-history characteristics may be more important for other
questions, like estimating species extinction risk (J. A. Hutchings et
al. 2012). For example, Blanchard, Maxwell, and Jennings (2007) used
detailed simulations of spatially-distributed fisheries to compare
survey designs. They found that statistical power depended on survey
design, temperature preferences and the degree of population patchiness.

An important related question, is the optimal allocation of sampling
effort in space versus time. In a theoretical investigation of this
question, Rhodes and Jonzen (2011) found that the optimal allocation of
sampling depended strongly on temporal and spatial autocorrelation. If
spatial population dynamics were highly correlated, then it was better
to sample more temporally, and vice versa. Our work supports this idea
as populations with strong temporal autocorrelation needed less years of
sampling (Fig. \ref{fig:correlates}). Morrison and Hik (2008) also
studied the optimal allocation of sampling effort in space versus time,
but used emprical data from a long-term survey of the collared pika
(\emph{Ochotona collaris}) found in the Yukon. They estimated long-term
growth rates among three subpopulations over a 10-year period. They
found that surveys less than 5 years may be misleading and that
extrapolating from one population to another, even when nearby
geographically, may be untenable.

Seavy and Reynolds (2007) asked whether statistical power was even a
useful framework for assessing long-term population trends. They used 24
years of census data on Red-tailed Tropicbirds
(\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50\%
decline over 10 years almost always resulting in high statistical power
(above 0.8). Therefore, they cautioned against only using power analyses
to design monitoring schemes and instead argued for metrics that would
increase precision. For example, Seavy and Reynolds (2007) suggest
improving randomization, reducing bias, and increases detection
probability when designing and evaluating monitoring programs. We agree
that power analyses should not be the only consideration when designing
monitoring schemes. However, unlike Seavy and Reynolds (2007), our
results indicate that longer than 10 years is often needed to achieve
high statistical power.

The design of monitoring programs should include calculations of
statistical power, the allocation of sampling in space versus time
(Rhodes and Jonzen 2011), and metrics to increase precision. Ideally, a
formal decision analysis to evaluate these different factors would be
conducted to design or assess any monitoring program (Hauser, Pople, and
Possingham 2006; McDonald-Madden et al. 2010). This type of formal
decision analysis would also include information on the costs of
monitoring. These costs include the actual costs of sampling (Brashares
and Sam 2005) and the ecological costs on inaction (Thompson et al.
2000).

@@

Our work has some limitations in determining the minimum time series
length required. First, \(T_{min}\) is particular to the specific
question of interest. An additional complication is that for our
empirical approach, the subsampling of the full time series allows for
estimates of power, but the individual subsamples are clearly not
independent of one another. Further, estimates of \(T_{min}\) depend on
chosen values of \(\alpha\) and \(\beta\) (Fig.
\ref{fig:min_time_vs_alpha_beta}). In an ideal setting, we would build a
specific population model parameterized for each population of interest
(McCain, Szewczyk, and Knight 2016). Then, model simulations could be
used to estimate the minimum time series required to address each
specific question of interest. Clearly, this is not always practical,
especially if conducting analyses for a wide array of species as we do
here. In addition, our statistical models suggest that \(T_{min}\) does
not correlate with any biological variables of interest, at least for
the question of linear regression (Fig.
\ref{fig:biological_correlates}). Therefore, it is not possible to use
these results to predict \(T_{min}\) for another population, even if the
population is of a species with a similar life-history to one in our
database.

@@

We use a database of 878 populations to determine the minimum time
series length required to detect population trends. This goes beyond
previous work that either focused on theoretical investigations or a
limited number of species. We show that to identify long-term changes in
abundance, 15-20 years of continuous monitoring are often required (Fig.
\ref{fig:min_time_dist}). In line with theoretical predictions
(Gerrodette 1987), we also show that \(T_{min}\) is strongly correlated
with the long-term population trend (i.e.~rate of increase), variability
in population size, and the temporal autocorrelation (Fig.
\ref{fig:correlates}).

Our work implies that for many populations, time series less than 15-20
years are probably not reliable for detecting population trends. This
result highlights the importance of long-term monitoring programs. From
both a scientific and management perspective estimates of \(T_{min}\)
are important. If a time series is too short, we lack statistical power
to reliably detect long-term population trends. In addition, a time
series that is too long may be a poor use of already limited funds
(Gerber, DeMaster, and Kareiva 1999). However, short time series data
can not be used as an excuse for inaction. For example, Thompson et al.
(2000) demonstrate that long-term studies may be required to detect
population trends for certain species. When a population trend is
detected, it may be too late for management action. In these situations,
the precautionary principle may be more appropriate (Thompson et al.
2000). Future work should examine other species, with a wider range of
life history characteristics. In addition, similar approaches can be
used to determine the minimum time series length required to address
additional questions of interest.

@@

\hypertarget{refs}{}
\hypertarget{ref-Blanchard2007}{}
Blanchard, Julia L, David L Maxwell, and Simon Jennings. 2007. ``Power
of monitoring surveys to detect abundance trends in depleted
populations: the effects of density-dependent habitat use, patchiness,
and climate change.'' \emph{ICES Journal of Marine Science} 65 (1):
111--20.

\hypertarget{ref-Bolker2008}{}
Bolker, Benjamin M. 2008. \emph{Ecological Models and Data in R}. 1st
ed. Princeton, New Jersey: Princeton University Press.

\hypertarget{ref-Brashares2005}{}
Brashares, Justin S, and Moses K Sam. 2005. ``How much is enough?
Estimating the minimum sampling required for effective monitoring of
African reserves.'' \emph{Biodiversity and Conservation} 14: 2709--22.
doi:\href{https://doi.org/10.1007/s10531-005-8404-z}{10.1007/s10531-005-8404-z}.

@@

\hypertarget{ref-McCain2016}{}
McCain, Christy Marie, Tim Szewczyk, and Kevin Bracy Knight. 2016.
``Population variability complicates the accurate detection of climate
change responses.'' \emph{Global Change Biology} 22 (6): 2081--93.
doi:\href{https://doi.org/10.1111/gcb.13211}{10.1111/gcb.13211}.

@@

\hypertarget{ref-Thompson2000}{}
Thompson, Paul M, Ben Wilson, Kate Grellier, and Philip S Hammond. 2000.
``Combining power analysis and population viability analysis to compare
traditional and precautionary approaches to conservation of coastal
cetaceans.'' \emph{Conservation Biology} 14 (5): 1253--63.

### 2017-10-01 writing/supp_mat_draft.Rmd
Version submitted to Biological Conservation and the second version to PeerJ preprints

June 22, 2017
 \end{center}

@@

We then perform a subsampling routine to estimate the minimum time required $T_{min}$ (similar to @Gerber1999). This is the same routine we used for the results in the main manuscript.
\begin{enumerate}
  \item We first extract all contiguous subsamples of the time series. This leads to 34 two-year subsamples, 33 two-year subsamples, and so forth until a single 35-year subsample.  
  \item For each subsample, we conduct linear regression and extract model coefficients and p-values.    \item We can call each set of subsamples, of the same length, a set. The fraction of subsamples within a set that show significant trends (significant slope coefficient $\alpha$ less than 0.05) is the statistical power. It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series. In other words, we are conducting a one-tailed test.
  \item We can then plot statistical power as a function of time series length (Fig. \ref{fig:empirical_approach_example}b). As expected, we can see that power increases with the more years that are sampled.
  \item  Then, we determine an appropriate level of statistical power that we find acceptable. Traditionally, this has been at 0.8, however, this is purely historical. Statistical power of 0.8 implies if a true trend is present, or there is a real change in abundance, then we will detect this trend 0.8 fraction of the time. 
  \item We then determine the minimum time series length ($T_{min}$) required to achieve that level of statistical power. Here, $T_{min}$ is the first point in Fig. \ref{fig:empirical_approach_example}b where following points are also above 0.8. In this example, $T_{min}$ is `r head(which(power_example>0.8),1)`. 
\end{enumerate}

@@

We simulate this model repetitively for different lengths of time series. Statistical power is the fraction of time series of a set length that showed significant trends in abundance, given a true trend in abundance. *How do I determine significance here? Through linear regression again, or do I look at non-zero estimation of the parameter values?*-->

>>>>

 \end{center}

@@

We then perform a subsampling routine to estimate the minimum time required $T_{min}$ (similar to @Gerber1999). This is the same routine we used for the results in the main manuscript.
\begin{enumerate}
  \item We first extract all contiguous subsamples of the time series. This leads to 34 two-year subsamples, 33 two-year subsamples, and so forth until a single 35-year subsample.  
  \item For each subsample, we conduct linear regression and extract model coefficients and p-values.    \item We can call each set of subsamples, of the same length, a set. The proportion of subsamples within a set that show significant trends (significant slope coefficient $\alpha$ less than 0.05) is the statistical power. It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series. In other words, we are conducting a one-tailed test.
  \item We can then plot statistical power as a function of time series length (Fig. \ref{fig:empirical_approach_example}b). As expected, we can see that power increases with the more years that are sampled.
  \item  Then, we determine an appropriate level of statistical power that we find acceptable. Traditionally, this has been at 0.8, however, this is purely historical. Statistical power of 0.8 implies if a true trend is present, or there is a real change in abundance, then we will detect this trend 0.8 proportion of the time. 
  \item We then determine the minimum time series length ($T_{min}$) required to achieve that level of statistical power. Here, $T_{min}$ is the first point in Fig. \ref{fig:empirical_approach_example}b where following points are also above 0.8. In this example, $T_{min}$ is `r head(which(power_example>0.8),1)`. 
\end{enumerate}

@@

We simulate this model repetitively for different lengths of time series. Statistical power is the proportion of time series of a set length that showed significant trends in abundance, given a true trend in abundance. *How do I determine significance here? Through linear regression again, or do I look at non-zero estimation of the parameter values?*-->

### 2017-10-01 writing/supp_mat_draft.tex
Version submitted to Biological Conservation and the second version to PeerJ preprints

\begin{enumerate}
  \item We first extract all contiguous subsamples of the time series. This leads to 34 two-year subsamples, 33 two-year subsamples, and so forth until a single 35-year subsample.  
  \item For each subsample, we conduct linear regression and extract model coefficients and p-values.    \item We can call each set of subsamples, of the same length, a set. The fraction of subsamples within a set that show significant trends (significant slope coefficient $\alpha$ less than 0.05) is the statistical power. It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series. In other words, we are conducting a one-tailed test.
  \item We can then plot statistical power as a function of time series length (Fig. \ref{fig:empirical_approach_example}b). As expected, we can see that power increases with the more years that are sampled.
  \item  Then, we determine an appropriate level of statistical power that we find acceptable. Traditionally, this has been at 0.8, however, this is purely historical. Statistical power of 0.8 implies if a true trend is present, or there is a real change in abundance, then we will detect this trend 0.8 fraction of the time. 
  \item We then determine the minimum time series length ($T_{min}$) required to achieve that level of statistical power. Here, $T_{min}$ is the first point in Fig. \ref{fig:empirical_approach_example}b where following points are also above 0.8. In this example, $T_{min}$ is 22. 
\end{enumerate}

>>>>

\begin{enumerate}
  \item We first extract all contiguous subsamples of the time series. This leads to 34 two-year subsamples, 33 two-year subsamples, and so forth until a single 35-year subsample.  
  \item For each subsample, we conduct linear regression and extract model coefficients and p-values.    \item We can call each set of subsamples, of the same length, a set. The proportion of subsamples within a set that show significant trends (significant slope coefficient $\alpha$ less than 0.05) is the statistical power. It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series. In other words, we are conducting a one-tailed test.
  \item We can then plot statistical power as a function of time series length (Fig. \ref{fig:empirical_approach_example}b). As expected, we can see that power increases with the more years that are sampled.
  \item  Then, we determine an appropriate level of statistical power that we find acceptable. Traditionally, this has been at 0.8, however, this is purely historical. Statistical power of 0.8 implies if a true trend is present, or there is a real change in abundance, then we will detect this trend 0.8 proportion of the time. 
  \item We then determine the minimum time series length ($T_{min}$) required to achieve that level of statistical power. Here, $T_{min}$ is the first point in Fig. \ref{fig:empirical_approach_example}b where following points are also above 0.8. In this example, $T_{min}$ is 22. 
\end{enumerate}
